Robertson and Cao Project

### 2024-04-02 products/manuscript/Manuscript.qmd
Transferred code from .R to .qmd, updated README
Updated README to reflect changes and transfers

# Introduction
## Background

Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings [1]. HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime [2, 3]. Most cases of HPV are asymptomatic and cleared over time, but certain infections tend to become persistent and can cause genital warts. Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells [4]. The HPV group contains over 200 related viruses, but only 12 types are considered “high-risk” due to the presence of oncogenes. These include HPV 16, 18, 31, 33, 35, 39, 45, 51, 52, 56, 58, and 59, though HPV-16 and HPV-18 are associated with the highest number of HPV-associated cancer cases. In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers [5,6]. In an effort to prevent HPV-associated cancers worldwide, two HPV vaccines were developed and made available in 2006. These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types [7]. In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 [4]. Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first [8]. In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime [9].

@@

Measuring HPV vaccine completion with “up-to-date” variables may be more robust than vaccination rate based on those who have completed the regime (2 or more shots), because not all teens included in the survey are at the age in which they would’ve completed the full regime. This means that younger teens who are up to date with 1+ shot will not be excluded from the analysis. We aim to measure determinants of socioeconomic status that are engrained into the survey questions. These factors include: family income, poverty status, income-to-poverty ratio, insurance status, insurance breaks, maternal education, living arrangement, and geographic mobility status. Additional demographic factors that are associated with healthcare access and may be examined include: race, ethnicity, language, facility in which the vaccine was administered, and whether the teen had completed a wellness exam between the ages of 11-12. Geographic distribution of the teens will be assessed by true state of residence. We would like to examine the data for correlations between HPV vaccination completion status and the aforementioned socioeconomic determinants, demographics, and geographic location. This serves as valuable information to determine differences in healthcare access and vaccination coverage for teenagers living in the U.S.
It is generally known that those of lower income and without insurance have limited healthcare access in the U.S. [10], but it will be interesting to see if this pattern is reflected at the state level. If so, policy should be examined as it may be related to healthcare access disparities. Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 [11]. More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic [12]. Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.

@@

![Figure 1. Income-poverty ratio distribution](/results/figures/income.png)

@@

{{< pagebreak >}} 
# Analysis

@@

# References
1.	https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length. 
2.	https://www.womenshealth.gov/a-z-topics/human-papillomavirus#:~:text=Genital%20HPV%20is%20the%20most,79%20million%20Americans%20have%20HPV.&text=It%20is%20so%20common%20that,some%20point%20in%20their%20lifetime.
3.	https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.
4.	HPV and Cancer. In: National Cancer Institute [Internet]. 1 Mar 2019 [cited 12 Mar 2024]. Available: https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
5.	https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer
6.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html#:~:text=Vaccine%20Composition,45%2C%2052%2C%20and%2058. 
7.	https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length. 
8.	https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html#:~:text=Two%20doses%20of%20HPV%20vaccine,third%20dose%20of%20HPV%20vaccine.
9.	https://www.cdc.gov/mmwr/volumes/72/wr/mm7234a3.htm#:~:text=In%202022%2C%2089.9%25%20of%20adolescents,HPV%20vaccination%20(HPV%20UTD).

10.  Sciences, N. A. of, Engineering, & Medicine, and. (2018, March 1). Factors that affect health-care utilization. Health-Care Utilization as a Proxy in Disability Determination. https://www.ncbi.nlm.nih.gov/books/NBK500097/
11.  Mansfield, L. N., Chung, R. J., Silva, S. G., Merwin, E. I., & Gonzalez-Guarda, R. M. (2022). Social determinants of human papillomavirus vaccine series completion among U.S. adolescents: A mixed-methods study. SSM - population health, 18, 101082. https://doi.org/10.1016/j.ssmph.2022.101082
12.  Casey, R. M., Akaba, H., Hyde, T. B, et al. (2024). Covid-19 pandemic and equity of global human papillomavirus vaccination: descriptive study of World Health Organization-Unicef vaccination coverage estimates. BMJ Medicine, 3, 000726. doi: 10.1136/bmjmed-2023-000726
13.  https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule
14.  https://www.paho.org/en/news/5-9-2023-paho-technical-advisory-group-recommends-countries-americas-use-single-dose-hpv

>>>>

# Introduction

## Background

Human Papillomavirus (HPV) is a double-stranded DNA virus that infects the cells of the skin and mucosal linings \[1\].
HPV is the most common sexually transmitted infection (STI) in the United States, where over 90% of sexually active males and 80% of sexually active females will likely become infected within their lifetime \[2, 3\].
Most cases of HPV are asymptomatic and cleared over time, but certain infections tend to become persistent and can cause genital warts.
Specific types of HPV are known to be “high-risk” or increase the likelihood of developing cancer in persistently infected cells \[4\].
The HPV group contains over 200 related viruses, but only 12 types are considered “high-risk” due to the presence of oncogenes.
These include HPV 16, 18, 31, 33, 35, 39, 45, 51, 52, 56, 58, and 59, though HPV-16 and HPV-18 are associated with the highest number of HPV-associated cancer cases.
In the U.S., there are 37,300 cases of HPV-associated cancer cases every year, including cases of cervical, anal, oropharyngeal (throat and mouth), vulvar, vaginal, and penile cancers \[5,6\].
In an effort to prevent HPV-associated cancers worldwide, two HPV vaccines were developed and made available in 2006.
These included a bivalent and quadrivalent vaccine that contained viral-like particles (VLPs) of the viral capsid protein (L1) from “high-risk” HPV types \[7\].
In recent years, the vaccine used in the U.S., Gardasil 9, has been expanded to contain HPV types 6, 11, 16, 18, 31, 33, 45, 52, and 58 \[4\].
Currently, the United States Centers for Disease Control and Prevention (CDC) recommends a two-dose Gardasil 9 regime to all teenagers, with a primary dose at age 11 to 12 and a second dose within 6 to 12 months of the first \[8\].
In 2022, the CDC estimated that 76.0% of teenagers aged 13 to 17 had received at least 1 dose of the HPV vaccine while only 62.6% of teenagers were up-to-date with their vaccination regime \[9\].

@@

Measuring HPV vaccine completion with “up-to-date” variables may be more robust than vaccination rate based on those who have completed the regime (2 or more shots), because not all teens included in the survey are at the age in which they would’ve completed the full regime.
This means that younger teens who are up to date with 1+ shot will not be excluded from the analysis.
We aim to measure determinants of socioeconomic status that are engrained into the survey questions.
These factors include: family income, poverty status, income-to-poverty ratio, insurance status, insurance breaks, maternal education, living arrangement, and geographic mobility status.
Additional demographic factors that are associated with healthcare access and may be examined include: race, ethnicity, language, facility in which the vaccine was administered, and whether the teen had completed a wellness exam between the ages of 11-12.
Geographic distribution of the teens will be assessed by true state of residence.
We would like to examine the data for correlations between HPV vaccination completion status and the aforementioned socioeconomic determinants, demographics, and geographic location.
This serves as valuable information to determine differences in healthcare access and vaccination coverage for teenagers living in the U.S.
It is generally known that those of lower income and without insurance have limited healthcare access in the U.S. \[10\], but it will be interesting to see if this pattern is reflected at the state level.
If so, policy should be examined as it may be related to healthcare access disparities.
Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 \[11\].
More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic \[12\].
Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.

@@

![Figure 1. Income-poverty ratio distribution](../results/figures/income.png)

@@

{{< pagebreak >}} \# Analysis

@@

1.  https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:\~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.

2.  https://www.womenshealth.gov/a-z-topics/human-papillomavirus#:\~:text=Genital%20HPV%20is%20the%20most,79%20million%20Americans%20have%20HPV.&text=It%20is%20so%20common%20that,some%20point%20in%20their%20lifetime.

3.  https://www.health.state.mn.us/data/mcrs/data/qfhpv.html#:\~:text=Human%20Papillomavirus%20(HPV)%20is%20a,with%20HPV%20in%20their%20lifetime.&text=Around%2050%20percent%20of%20HPV,HPV%2C%20which%20can%20cause%20cancer.

4.  HPV and Cancer.
    In: National Cancer Institute \[Internet\].
    1 Mar 2019 \[cited 12 Mar 2024\].
    Available: https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer

5.  https://www.cancer.gov/about-cancer/causes-prevention/risk/infectious-agents/hpv-and-cancer

6.  https://www.cdc.gov/vaccines/vpd/hpv/hcp/vaccines.html#:\~:text=Vaccine%20Composition,45%2C%2052%2C%20and%2058.

7.  https://www.who.int/teams/health-product-policy-and-standards/standards-and-specifications/vaccine-standardization/human-papillomavirus#:\~:text=Human%20papillomavirus%20(HPV)%20is%20a,approximately%208%2Dkb%20in%20length.

8.  https://www.cdc.gov/vaccines/vpd/hpv/hcp/recommendations.html#:\~:text=Two%20doses%20of%20HPV%20vaccine,third%20dose%20of%20HPV%20vaccine.

9.  https://www.cdc.gov/mmwr/volumes/72/wr/mm7234a3.htm#:\~:text=In%202022%2C%2089.9%25%20of%20adolescents,HPV%20vaccination%20(HPV%20UTD).

10. Sciences, N. A.
    of, Engineering, & Medicine, and.
    (2018, March 1).
    Factors that affect health-care utilization.
    Health-Care Utilization as a Proxy in Disability Determination.
    https://www.ncbi.nlm.nih.gov/books/NBK500097/

11. Mansfield, L. N., Chung, R. J., Silva, S. G., Merwin, E. I., & Gonzalez-Guarda, R. M.
    (2022).
    Social determinants of human papillomavirus vaccine series completion among U.S. adolescents: A mixed-methods study.
    SSM - population health, 18, 101082.
    https://doi.org/10.1016/j.ssmph.2022.101082

12. Casey, R. M., Akaba, H., Hyde, T. B, et al. (2024).
    Covid-19 pandemic and equity of global human papillomavirus vaccination: descriptive study of World Health Organization-Unicef vaccination coverage estimates.
    BMJ Medicine, 3, 000726.
    doi: 10.1136/bmjmed-2023-000726

13. https://www.who.int/news/item/20-12-2022-WHO-updates-recommendations-on-HPV-vaccination-schedule

14. https://www.paho.org/en/news/5-9-2023-paho-technical-advisory-group-recommends-countries-americas-use-single-dose-hpv
